HyperLogLog: Exponentially Bad in Adversarial Settings

### 2020-08-19 docs/HLL_exp_bad/main.tex
Update from report

\documentclass[11pt]{article}
\usepackage{graphicx}
\usepackage{array}
\usepackage{url}
\usepackage{hyperref}
\usepackage{xspace}
\usepackage{a4wide}

%% TODO NOTES

\usepackage{xcolor}
\definecolor{oxygenorange}{HTML}{FFDD00}
\usepackage[color=oxygenorange]{todonotes}
\newcommand{\mathilde}[1]{\todo[inline]{\textbf{Mathilde:} #1}\xspace}
\newcommand{\kenny}[1]{\todo[inline]{\textbf{Kenny:} #1}\xspace}

@@

Very recently, Reviriego and Ting~\cite{hllvuln} initiated work in this direction, showing that in specific attack settings, the cardinality estimate made by HLL could be modified through input selection. In this paper, we present a complete analysis of HLL in adversarial setting, considering more realistic attack scenarios and giving more powerful attacks than~\cite{hllvuln}. In particular, we show that with only modest knowledge of the HLL internals and a moderate amount of computation, an adversary with sufficient flexibility in the choice of inputs can make the HLL cardinality estimate exponentially smaller than the true cardinality (e.g.\ constant instead of $O(2^t)$ where $t$ is th number of bits of flexibility in the input).

The rest of the paper is organised as follows. Section~\ref{sec:overview} gives an overview of the HyperLogLog algorithm. Section~\ref{sec:attacks} presents our different adversarial models. There, we also give attacks and evaluate their impacts under the different adversarial models. Section~\ref{sec:conclusions} provides our conclusions and some ideas for future work.

\section{Overview of HLL}\label{sec:overview}
\subsection{The Algorithm}
HLL~\cite{hll} is a streaming algorithm based on the key observation that, for a stream of randomly distributed values represented as bit-strings of some fixed length, if we observe a value with the maximum of $k$ leading zero-bits, then the cardinality of the stream (i.e.\ the number of distinct values it contains) is likely to be on the order of $2^{k+1}$. In practice, to ensure a uniform distribution, the stream values are first processed by a hash function $h$ before the leading bits are inspected. So to estimate the cardinality of a stream, we need only store a representation of $k$, the length of the largest observed string of leading zero-bits for numbers in the stream. This provides a very compact mechanism for estimating the stream cardinality -- if the true cardinality is $N$, then just $\log\log(N)$ bits are needed to store the estimate.

This simple approach suffers from large variance in the estimation. In order to improve the estimate, we can use many estimators in parallel instead of one, and average the results. Each estimate is stored in its own bucket. The collection of buckets is referred to as the HLL \emph{sketch}. To map a value $x$ to one of $2^n$ buckets, \cite{loglog} suggests using the first $n$ bits of $h(x)$ as a bucket index, and then computing the longest sequence of leading zero bits on the remaining $\ell$ bits of $h(x)$ (so that the output length of $h(\cdot)$ is $n+\ell$). Estimators from all buckets are averaged using the harmonic mean and scaled by a constant $\alpha$, where $\alpha$ is empirically computed to correct a systematic multiplicative bias. Small and big cardinalities are handled differently to correct eventual distortions, as shown in Fig. \ref{fig:hll}. When the estimate is comparatively small with respect to $2^n$, the Hit Counting algorithm of Whang et al.~\cite{hitcounting} which looks at the number of empty bins is used to provide the estimator. A similar correction is applied for large cardinalities in order to take into account the likelihood of hashing collisions. As result, HLL is able to estimate cardinalities greater than $10^9$ with a typical standard error of 2\%, using only 1.5 kB of memory~\cite{hll}.

An interesting property of HLL is that it supports merging in a lossless way. To combine two buckets at index $i$ in two different HLL instances, take the maximum of the two bucket entries and assign that to the matching bucket $i$ in the merged HLL sketch. Such a simple set union operation allows easy parallelisation of operations among multiple machines independently, provided they use the same hash function and the same number of buckets.

\begin{figure}[ht]
\centering
\fbox{
\begin{minipage}{40em}
\begin{flushleft}
\textit{Let $h : D \rightarrow \{0, 1\}^{32}$ hash data from $\mathcal{D}$ to binary $32$–bit words.\\
Let $\rho(s)$ be the position of the leftmost $1$-bit of $s$: e.g., $\rho(1...) = 1, \rho(0001...) = 4, \rho(0^K) = K + 1.$\\}
\textbf{define} $\alpha_{16} = 0.673$; $\alpha_{32} = 0.697$; $\alpha_{64} = 0.709$; $\alpha_{m} = 0.7213/(1+ 1.079/m)$ for $m \ge 128$;\\
\vspace{0.5em}
\textbf{Program} HyperLogLog (\textbf{input} $\mathcal{M}$ : multiset of items from domain $\mathcal{D}$).\\
\textbf{assume} $m = 2^n$ with $n \in [4..16].$\\
\textbf{initialize} a collection of $m$ registers, $M[1], ..., M[m]$, to $0$;\\
\vspace{0.5em}
\textbf{for} $v \in \mathcal{M}$ \textbf{do}\\
    \hspace{2em} \textbf{set} $x := h(v);$\\
    \hspace{2em} \textbf{set} $j = 1 + <x_1x_2...x_{n}>_2$; \hfill \textit{\{the binary address determined by the first $b$ bits of $x$\}}\\
    \hspace{2em} \textbf{set} $w := x_{n+1}x_{n+2}...$;\\
   \hspace{2em} \textbf{set} $M[j] :=$ max$(M[j], \rho(w))$;\\
\vspace{0.2em}
\textbf{compute} $E := \alpha_mm^2\cdot(\sum_{j=1}^m2^{-M[j]})^{-1}$; \hfill \textit{\{the “raw” HyperLogLog estimate\}}\\
\vspace{0.2em}
\textbf{if} $E \le \frac{5}{2}m$ \textbf{then}\\
    \hspace{2em}  \textbf{let} $V$ be the number of registers equal to $0$;\\
    \hspace{2em}  \textbf{if} $V \neq 0$ \textbf{then set} $E^*:= m $log$(m/V)$ \textbf{else set} $E^*:= E$; \hfill \textit{\{small range correction\}}\\
\textbf{if} $E \le \frac{1}{30}2^{32}$ \textbf{then}\\
    \hspace{2em} \textbf{set} $E^*:= E$; \hfill \textit{\{intermediate range—no correction\}}\\
\textbf{if} $E > \frac{1}{30}2^{32}$ \textbf{then}\\
    \hspace{2em}  \textbf{set} $E^*:= -2^{32}$log$(1 - E/2^{32})$; \hfill \textit{\{large range correction\}}\\
\textbf{return} \textit{cardinality estimate $E^*$ with typical relative error} $\pm1.04/\sqrt{m}$.
\end{flushleft}
\end{minipage}
}
\caption{The HyperLogLog Program}
\label{fig:hll}
\end{figure}

@@

Regardless of the sensitivity of the above examples, none of the relevant papers are clear about their threat model, especially for the HLL component.  They are all potentially vulnerable to the attacks we present below.
%By not assessing the risks involved with the presence of an adversary targeting the HLL sketch, they may allow an attack like the one we present, breaking down the whole solution.

\subsection{Related Work}

@@

In~\cite{hllvuln}, Reviriego and Ting exploit a vulnerability of HLL to manipulate the cardinality estimate, leading to a five-fold reduction in the estimate compared to the true cardinality. They use an attack model that we consider to be quite artificial, since it gives the attacker the ability to test whether inserting an item \emph{would} increase the cardinality estimate, without actually having to insert the item. See further discussion in Section~\ref{sec:attacks}.

@@

Throughout, we assume the adversary has the ability to vary the contents of free fields in the input strings which will be inserted into the stream of values. For example it can manipulate IPID and other header fields in the context of IP packets. This enables the adversary to control the output of hash function $h$ and flexibly choose which items will be inserted into the HLL sketch, and in which buckets.

\subsection{Adversarial Setting}
We model four distinct adversarial scenarios based on the adversary's knowledge and capabilities. For all of them, we assume that the sketch already contains some existing data from honest users to simulate a real-life setting. Among the scenarios, the \textit{insider} setup of \cite{cardestprivacy} that we present as S4 makes the strongest assumptions about the adversarial capabilities since it assumes that the attacker has a perfect view on the sketch. %combines the knowledge of adversaries from S1 and S2.
%\kenny{Not really! Because S1 is actually quite unrealistic, as we will argue below.}

\begin{itemize}
\item[S1:] The adversary does not know the details of the target HLL implementation but can access a copy of the HLL instance as a user. He can use the API to insert elements into this shadow sketch, get its cardinality estimate, and reset it to its native empty state. The attacker does not know the state of the targeted HLL sketch, meaning that he does not have the list of elements previously inserted into it by other users, nor can he obtain its cardinality estimate.
\item[S2:] The adversary has access to the details of the HLL implementation, i.e., the adversary knows the number of buckets $2^n$ and the hash function $h$ in use, but is otherwise ``blind''. In particular, an S2 adversary does not have access to the values of the HLL cardinality estimate at any point in its attack. This kind of attack scenario is appropriate in the context of the port scanning attack of~\cite{portscanhll}.
\item[S3:] The adversary has access to the details of the HLL implementation, i.e., the adversary knows the number of buckets $2^n$ and the hash function $h$ in use. Additionally, it can access and interact with the sketch via an API provided by the sketch owner, allowing it to insert items into the HLL sketch and ask for the HLL's cardinality estimate at any point in time.
\item[S4:] The adversary has direct access to the HLL in use and all its internals, i.e., the number of buckets, the hash function $h$, and also the contents of each bucket in the HLL sketch \emph{at one specific point in time}, i.e.\ it is given a \emph{snapshot} of the sketch. It then causes chosen inputs to be inserted into the HLL sketch. An S4 adversary can compute a lower bound on the HLL cardinality estimate based on its snapshot and whatever it causes to be inserted into the HLL sketch during its attack (this is irrespective of whether its attack proceeds with input from other users being added to the HLL sketch in parallel with its attack or not; if it knows that there is no other input, then it can compute the \emph{exact} value of the HLL cardinality estimate at every point in its attack instead of a lower bound).
\end{itemize}

@@

We start by highlighting the fact that S1 seems unrealistic in practice. In a networking application, the authors of~\cite{hllvuln} argue that the assumptions of S1 could be fulfilled if the attacker learns which machine is used to do the monitoring, buys the same device, and uses it as the shadow device. It is possible that the provider uses a different HLL salt or key per device. This would make any pre-computation with the shadow device irrelevant and considerably weaken the attack presented in Section 3.2 of \cite{hllvuln}. If, on the other hand, we assume that the devices are perfect copies of one another, then we show that it is actually possible for the attacker to predict the behaviour of the HLL sketch as elements are inserted into it, and effectively elevate itself to being an S2 adversary, without actually knowing the hash function $h$ in use. In detail, the attacker can recover some bits of interest in the hash values of targeted items, and infer the knowledge given to the adversary under S2. To do so, the attacker can insert the item $x$ on an empty sketch and ask for the resulting cardinality, denoted $c$. Note that inserting $x$ only touches one bucket of the HLL sketch; let that bucket have index $i$ (which is unknown to the attacker) and value $c_i$ after the insertion.  From the formula used to compute the HLL sketch cardinality estimate $c$, we get the equation:
\begin{eqnarray*}
   \log_2(c)-\log_2(\alpha 2^{2n} - c2^n + c) &=& c_i
\end{eqnarray*}
which enables the adversary to learn the value $c_i$ from $c$. (Here $\alpha$ is a constant, and $2^n$ is known since the provider must explicitly tell the customer the expected error of the estimation, hence the number of buckets.) Now $c_i-1$ is (an approximation of) the number of leading zeros in the $l$ rightmost bits of $h(x)$. This is precisely the information extracted from $h(x)$ and used by the adversary in his attack in scenario S2, as we detail in Section~\ref{sec:attacks}. %the value of the updated estimator of the bucket $x$ was mapped to, and recover some bits of $h(x)$.
To summarize, we do not consider scenario S1 typical of HLL deployments, and even if such a setup may be possible under certain circumstances, the attacker can infer the information used in our attack in scenario S2.

%\mathilde{Do you think that we should mention the idea behind the attack of \cite{hllvuln} Section 3.2 ? As they explained in their emails "The attacker can then test and find elements that do not increase the HLL estimate on the shadow HLL and thus will unlikely increase the cardinality of the target HLL when inserted."}

Furthermore, we show that although~\cite{cardestprivacy} presents S3 (external adversary) and the stronger version of our S4 (insider adversary) as two separate scenarios, an S3 adversary can easily infer the knowledge given to our S4 adversary in its snapshot. Indeed, an adversary can easily recover the contents of each bucket in the sketch simply by analysing which inputs increase the cardinality estimate. To find the estimate held in a targeted bucket, the attacker gradually inserts items $x$ with an incrementing number of leading zeros in the $\ell$ rightmost bits of $h(x)$ whilst holding the $n$ bits determining the bucket index constant, until the cardinality estimate of the sketch increases. Once an increase occurs, the attacker knows that the maximum number of leading zeros observed in that bucket was just updated to a value that can be recovered by looking at the last inserted item. This assumes the attacker has sufficient flexibility in its choice of inputs $x$ so that the relevant bit conditions on $h(x)$ can be forced. The attacker can determine all bucket contents with, in the worst case, $O(2^{n + \ell})$ pre-computation and $O(\ell \cdot 2^{n})$ insert and cardinality estimate queries, and needs about $n + \ell$ bits of flexibility assuming $h$ behaves like a random function. The pre-computation and query complexity required can be much less if the HLL sketch is only moderately loaded.

A version of this attack can be carried out to elevate an external adversary to a full-strength insider adversary as per~\cite{cardestprivacy} (instead of our slightly weaker S4 adversary), since it can simply be done repeatedly each time the external adversary wants to know the complete internals of the HLL sketch. Obvious optimisations can be carried out to reduce the query complexity each time the attack is run (since the bucket contents can only increase over time, some queries are redundant; moreover, the pre-computation is only needed once and is moderate given typical HLL parameters, e.g.\ $n+\ell = 32$).

Since the assumptions of S1 appear to us to be unrealistic and since an attacker can escalate from S3 to S4, we focus on scenarios S2 and S4 in what follows. We first present an attack in scenario S2, providing much stronger results than the corresponding attack of \cite{hllvuln} (in their equivalent scenario M1). We then use the additional information the adversary has access to in scenario S4 to make the attack even more powerful.

\subsection{Manipulating the HLL Cardinality Estimate}

\noindent\paragraph{Scenario S2} The adversary knows $h$ so it can adjust fields in the item $x$ it is trying to insert and keep only those for which the $\ell$ rightmost bits of $h(x)$ start with a one bit. In this case, the number of leading zero bits computed by the estimator of the destination bucket is always zero, so the estimator of this bucket is unlikely to be updated and the averaged approximate cardinality is likely to be left unchanged. For example, in the scenario of \cite{portscanhll}, the adversary could change the source port of its IP packet until it is satisfied with its hash value. Given sufficient flexibility in the input $x$, the adversary can insert many items into the HLL sketch without increasing the cardinality markedly, by repeatedly sampling from the input domain.

\noindent\paragraph{Scenario S4} The adversary can adopt the same strategy as for S2, but make the attack more efficient and reduce the sampling requirements with the additional information available in this scenario (namely, the count values per bucket in the HLL sketch). So, it can allow some leading zero bits in the $\ell$ rightmost bits of $h(x)$ so long as this does not increase the current bucket estimate. In other words, if $h(x)$ is mapped to the $i$-th bucket with estimated cardinality $2^{c_i + 1}$ (because a value with $c_i$ leading zeroes was previously observed for this bucket), then, instead of the strict restriction of having a one bit in the first position in the $\ell$ rightmost bits of $h(x)$, the adversary is satisfied when there is a one bit in any of the first $c_i+1$ positions in this substring. In case the adversary hits an empty bucket (for which the estimator is 0), it must skip that bucket completely. Under these conditions, the estimators in the buckets are never updated, and there is no effect on the cardinality estimate made by HLL.  A special case arises when the HLL is completely empty at the beginning of the attack; here the adversary is forced to increase the estimate in at least one bucket, but with sampling this can obviously be done in such a way as to increase the HLL estimate to $2$ in a single bucket, leaving all the other buckets untouched.

\subsection{Impact}
We now evaluate the impact of the above attacks on the HLL cardinality estimate. We assume that $h$ is a ``good" hash function, so that each of the bits of $h(x)$ can be regarded as being independent and uniformly random.

In scenario S2, the adversary needs to craft two distinct candidates for $x$ on average to obtain a one bit at the first position in the $\ell$ rightmost bits of $h(x)$, thus successfully performing the attack for a single insertion. Assuming the adversary has $t$ bits of flexibility in its choice of inputs, the adversary can expect to ``hide''  $2^{t-1}$ distinct items without increasing the HLL cardinality estimate at all, under the assumption that all buckets in the HLL sketch already hold a value $c_i \ge 1$. On the other hand, if some buckets are empty (but the adversary does not know which ones in the blind setting of S2), then we may expect a moderate increase in the HLL cardinality estimator. In the worst case, when all buckets are initially empty, the cardinality estimate will increase from 0 to at most $\alpha2^{n+1}$ (regardless of the number of items the adversary inserts). Notice that in all cases, provided $t$ is large enough, the adversary can keep the cardinality estimate linear in the total number of buckets ($2^n$) while the expected value of the true cardinality, $2^{t-1}$, can be made as large as the adversary pleases.

>>>>

\documentclass[11pt]{article}
\usepackage{graphicx, amsmath, amsthm, amssymb, setspace, array, url, hyperref, xspace}
\usepackage{a4wide}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}

@@

Very recently, Reviriego and Ting~\cite{hllvuln} initiated work in this direction, showing that in specific attack settings, the cardinality estimate made by HLL could be modified through input selection. In this paper, we present a complete analysis of HLL in adversarial setting, considering more realistic attack scenarios and giving more powerful attacks than~\cite{hllvuln}. In particular, we show that with only modest knowledge of the HLL internals and a moderate amount of computation, an adversary with sufficient flexibility in the choice of inputs can make the HLL cardinality estimate exponentially smaller than the true cardinality (e.g.\ constant instead of $O(2^t)$ where $t$ is the number of bits of flexibility in the input).

The rest of the paper is organised as follows. Section~\ref{sec:overview} gives an overview of the HyperLogLog algorithm. A brief summary of the related works can be found in Section \ref{sec:related}. Section~\ref{sec:attacks} presents our different adversarial models. There, we also give attacks and evaluate their impacts under the different adversarial models. The experimental results are discussed in Section~\ref{sec:exp} and Section~\ref{sec:conclusions} provides our conclusions and some ideas for future work.

\section{Overview of HLL}\label{sec:overview}
\subsection{The Algorithm}
HLL~\cite{hll} is a streaming algorithm based on the key observation that, for a stream of randomly distributed values represented as bit-strings of some fixed length, if we observe a value with the maximum of $k$ leading zero-bits, then the cardinality of the stream (i.e.\ the number of distinct values it contains) is likely to be on the order of $2^{k+1}$. In practice, to ensure a uniform distribution, the stream values are first processed by a hash function $h$ before the leading bits are inspected. So to estimate the cardinality of a stream, we need only to store a representation of $k$, the length of the largest observed string of leading zero-bits for numbers in the stream. This provides a very compact mechanism for estimating the stream cardinality -- if the true cardinality is $N$, then just $\log\log(N)$ bits are needed to store the estimate.

This simple approach suffers from large variance in the estimation. In order to improve the estimate, we can use many estimators in parallel instead of one, and average the results. Each estimate is stored in its own register, that we call a bucket. The collection of buckets is referred to as the HLL \emph{sketch}. To map a value $x$ to one of $2^n$ buckets, \cite{loglog} suggests using the first $n$ bits of $h(x)$ as a bucket index, and then computing the longest sequence of leading zero bits on the remaining $\ell$ bits of $h(x)$. The output length of $h(\cdot)$ is thus $n+\ell$, fixed to 32 in~\cite{hll}. Estimators from all buckets are averaged using the harmonic mean and scaled by a constant $\alpha$, where $\alpha$ is empirically computed to correct a systematic multiplicative bias.

Relatively small and big cardinalities are handled separately to avoid eventual distortions, as shown in Fig.~\ref{fig:hll}. When the raw cardinality estimate is below $\frac{5}{2}\times 2^n$, the Hit Counting algorithm of Whang et al.~\cite{hitcounting} is used to produce the final cardinality estimator, looking at the number of empty buckets (whose register is 0) instead of the number of leading zeros in the stream. A similar correction is applied for large cardinalities in order to take into account the likelihood of hashing collisions. Some works building on top of~\cite{hll} replace the Hit counting algorithm by other schemes and use a hash function with a larger output domain to reduce the chances of collision and thus get rid of the need for large cardinality range correction. For example, \cite{hllpratice} uses a 64-bit hash function and Linear Counting along with bias correction in the small cardinality range. In this report we target the original paper's implementation reported in Fig.~\ref{fig:hll} and leave for future work the analysis of its variations.

As result, HLL is able to estimate cardinalities greater than $10^9$ with a typical standard error of 2\%, using only 1.5 kB of memory~\cite{hll}.

An interesting property of HLL is that it supports merging in a lossless way. To combine two buckets at index $i$ in two different HLL instances, one can take the maximum of the two bucket entries and assign that to the matching bucket $i$ in the merged HLL sketch. Such a simple set union operation allows easy parallelisation of operations among multiple machines independently, provided they use the same hash function and the same number of buckets.

\begin{figure}[ht]
\centering
     \makebox[\linewidth]{
\fbox{
\begin{minipage}{40em}
\begin{flushleft}
\textit{Let $h : D \rightarrow \{0, 1\}^{32}$ hash data from $\mathcal{D}$ to binary $32$–bit words.\\
Let $\rho(s)$ be the position of the leftmost $1$-bit of $s$: e.g., $\rho(1...) = 1, \rho(0001...) = 4, \rho(0^K) = K + 1.$\\}
\textbf{define} $m = 2^n$ with $n \in [4..16].$\\
\textbf{define} $\alpha_{16} = 0.673$; $\alpha_{32} = 0.697$; $\alpha_{64} = 0.709$; $\alpha_{m} = 0.7213/(1+ 1.079/m)$ for $m \ge 128$;\\
\vspace{0.5em}
\textbf{Algorithm} HyperLogLog (\textbf{input} $\mathcal{M}$ : multiset of items from domain $\mathcal{D}$).\\
\textbf{initialize} a collection of $m$ registers, $M[1], ..., M[m]$, to $0$;\\
\vspace{0.5em}
\textbf{for} $v \in \mathcal{M}$ \textbf{do}\\
    \hspace{2em} \textbf{set} $x := h(v);$\\
    \hspace{2em} \textbf{set} $j = 1 + <x_1x_2...x_{n}>_2$; \hfill \textit{\{the binary address determined by the first $n$ bits of $x$\}}\\
    \hspace{2em} \textbf{set} $w := x_{n+1}x_{n+2}...$;\\
   \hspace{2em} \textbf{set} $M[j] :=$ max$(M[j], \rho(w))$;\\
\vspace{0.2em}
\textbf{compute} $E := \alpha_mm^2\cdot(\sum_{j=1}^m2^{-M[j]})^{-1}$; \hfill \textit{\{the “raw” HyperLogLog estimate\}}\\
\vspace{0.2em}
\textbf{if} $E \le \frac{5}{2}m$ \textbf{then}\\
    \hspace{2em}  \textbf{let} $V$ be the number of registers equal to $0$;\\
    \hspace{2em}  \textbf{if} $V \neq 0$ \textbf{then set} $E^*:= m $ln$(m/V)$ \textbf{else set} $E^*:= E$; \hfill \textit{\{small range correction\}}\\
\textbf{if} $E \le \frac{1}{30}2^{32}$ \textbf{then}\\
    \hspace{2em} \textbf{set} $E^*:= E$; \hfill \textit{\{intermediate range—no correction\}}\\
\textbf{if} $E > \frac{1}{30}2^{32}$ \textbf{then}\\
    \hspace{2em}  \textbf{set} $E^*:= -2^{32}$ln$(1 - E/2^{32})$; \hfill \textit{\{large range correction\}}\\
\textbf{return} \textit{cardinality estimate $E^*$ with typical relative error} $\pm1.04/\sqrt{m}$.
\end{flushleft}
\end{minipage}
}}
\caption{The HyperLogLog Algorithm from~\cite{hll}}
\label{fig:hll}
\end{figure}

@@

Regardless of the sensitivity of the above examples, none of the relevant papers are clear about their threat model, especially for the HLL component.  They are all potentially vulnerable to the attacks we present below.

\section{Related Work}\label{sec:related}

@@

In~\cite{hllvuln}, Reviriego and Ting exploit a vulnerability of HLL to manipulate the cardinality estimate, leading to a five-fold reduction in the estimate compared to the true cardinality. Their attack model can be considered to be quite artificial, since it gives the attacker the ability to test whether inserting an item \emph{would} increase the cardinality estimate, without actually having to insert the item. After some clarifications obtained from personal correspondence with the authors, we finally agree that such setup might be possible under certain circumstances. See further discussion in Section~\ref{sec:attacks}.

@@

Throughout, we assume the adversary has the ability to vary the contents of free fields in the input strings which will be inserted into the stream of values. For example it can manipulate IPID and other header fields in the context of IP packets. This enables the adversary to control the output of hash function $h$ and flexibly choose which items will be inserted into the HLL sketch.

\subsection{Adversarial Setting}\label{sec:set}
We model four distinct adversarial scenarios based on the adversary's knowledge and capabilities. To simulate real-life settings, we consider both cases when the sketch is shared and receives inputs from other honest users but also when inputs are under adversarial control only.
It is safe to assume that the adversary knows the setting being used. 
Those two settings differ mainly in the fact that the adversary is attacking a sketch that already contains other users' data in the former option, but is empty in the latter one.

Among the scenarios, the \textit{insider} setup of~\cite{cardestprivacy} that we present as S4 makes the strongest assumptions about the adversarial capabilities since it assumes that the attacker has a perfect view on the sketch.

\begin{itemize}
\item[S1:] The adversary does not know the details of the target HLL implementation but can access a copy of the HLL instance as a user. He can use the API to insert elements into this shadow sketch, get its cardinality estimate, and reset it to its native empty state. The attacker does not know the state of the targeted HLL sketch, meaning that he does not have the list of elements previously inserted into it by other users.
\item[S2:] The adversary has access to the details of the HLL implementation, i.e., the adversary knows the number of buckets $2^n$ and the hash function $h$ in use, but is otherwise ``blind''. In particular, an S2 adversary does not have access to the values of the HLL cardinality estimate at any point in its attack. This kind of attack scenario is appropriate in the context of the port scanning attack of~\cite{portscanhll}.
\item[S3:] The adversary has access to the details of the HLL implementation, i.e., the adversary knows the number of buckets $2^n$ and the hash function $h$ in use. Additionally, it can access and interact with the sketch via an API provided by the sketch owner, allowing it to insert items into the HLL sketch and ask for the HLL's cardinality estimate at any point in time.
\item[S4:] The adversary has direct access to the HLL in use and all its internals, i.e., the number of buckets, the hash function $h$, and also the contents of each bucket in the HLL sketch \emph{at one specific point in time}, i.e.\ it is given a \emph{snapshot} of the sketch. It then causes chosen inputs to be inserted into the HLL sketch. An S4 adversary can compute a lower bound on the HLL cardinality estimate based on its snapshot and whatever it causes to be inserted into the HLL sketch during its attack (this is irrespective of whether its attack proceeds with input from other users being added to the HLL sketch in parallel with its attack or not; if it knows that there is no other input, then it can compute the \emph{exact} value of the HLL cardinality estimate at every point in its attack instead of a lower bound).
\end{itemize}

@@

We start by highlighting the fact that we do not consider scenario S1 typical of HLL deployments. In a networking application, the authors of~\cite{hllvuln} argue that the assumptions of S1 could be fulfilled if the attacker learns which machine is used to do the monitoring, buys the same, and uses it as the shadow device. It is possible that the provider uses a different salt or key in the hash per device. This would make any pre-computation on the items using the shadow device irrelevant, and considerably weaken the attack presented in Section 3.2 of~\cite{hllvuln}. Hence, we assume for the rest of this work that shadow and targeted devices have the same internals and security parameters. If the HLL device receives inputs from several users, the attack from Section 3.2 of~\cite{hllvuln} is the easiest to carry out. It works as follow: "after each insertion [in the shadow device], elements that do not increase the cardinality estimate are retained". Iterating through the process several times refines the items' filtering before inserting them into the targeted device. We explore in this work the setup where inputs are coming only from the adversary, and propose an attack that provides better results than the one we just presented.

Furthermore, we show that although~\cite{cardestprivacy} presents S3 (external adversary) and the stronger version of our S4 (insider adversary) as two separate scenarios, an S3 adversary can easily infer the knowledge given to our S4 adversary in its snapshot. Indeed, an adversary can easily recover the contents of each bucket in the sketch simply by analysing which inputs increase the cardinality estimate. To find the estimate held in a targeted bucket, the attacker gradually inserts items $x$ with an incrementing number of leading zeros in the $\ell$ rightmost bits of $h(x)$ whilst holding the $n$ bits determining the bucket index constant, until the cardinality estimate of the sketch increases. Once an increase occurs, the attacker knows that the maximum number of leading zeros observed in that bucket was just updated to a value that can be recovered by looking at the last inserted item. This assumes the attacker has sufficient flexibility in its choice of inputs $x$ so that the relevant bit conditions on $h(x)$ can be forced. The attacker can determine all bucket contents with, in the worst case, $O(2^{n + \ell})$ pre-computation and $O(\ell \cdot 2^{n})$ insert and cardinality estimate queries, and needs about $n + \ell$ bits of flexibility assuming $h$ behaves like a random function. The pre-computation and query complexity required can be much less if the HLL sketch is only moderately loaded. A version of this attack can be carried out to elevate an external adversary to a full-strength insider adversary as per~\cite{cardestprivacy} (instead of our slightly weaker S4 adversary), since it can simply be done repeatedly each time the external adversary wants to know the complete internals of the HLL sketch. Obvious optimisations can be carried out to reduce the query complexity each time the attack is run (since the bucket contents can only increase over time, some queries are redundant; moreover, the pre-computation is only needed once and is moderate given typical HLL parameters, e.g.\ $n+\ell = 32$).

Since an attacker can escalate from S3 to S4, we focus on scenarios S1, S2 and S4 in what follows. We first present an attack in scenario S1, providing much stronger results than the corresponding attack of~\cite{hllvuln} (in their equivalent scenario M2). We then use the additional information the adversary has access to in scenarios S2 and S4 to make the attack even more powerful.

\subsection{Manipulating the HLL Cardinality Estimate}
\noindent\paragraph{Scenario S1} The goal of the attacker is to infer some information about targeted items' hash values, precisely their bucket mappings, without knowing neither $h$ nor the salt in use (if any). With this knowledge, the adversary is able to retain items that are mapped to a subset of $b$ buckets out of the $2^n$, and minimize the resulting estimated cardinality. To achieve this, the attacker can take advantage of the Hit Counting algorithm initial phase to recover the number of non-empty buckets (whose register is not 0) after inserting few items. Indeed, in the small cardinality range, the cardinality estimator E* is computed as E*$=2^n$ln$(2^n/V)$ with $V$ the number of empty buckets. For typical HLL parameters, e.g.\ $n = 8$, the exact value of $V$, hence $b$, can be recovered from E*. Moreover, when inserting an item in the small cardinality range, the adversary will see a increase in the cardinality estimate if and only if $V$ is modified. This means that the number of empty buckets decreased by 1, therefore that the freshly inserted item is mapped to a bucket that was not hit before.

Thus, if the adversary blindly picks the first $b$ buckets to be hit as its targeted buckets, he can then discard all items whose insertion causes the cardinality estimator to increase from $b+k$ to $b+k+1$ for $k \in \mathbb{N}$. In other words, when $b$ buckets are already hit and the insertion of an item causes the cardinality estimator to increase, the attacker can discard this item as he knows that a new bucket (other than one of the $b$ buckets to be filled) switched to a non-empty state. The adversary can keep inserting items and discarding some while in the small cardinality range.

As soon as the cardinality estimator exceeds $\frac{5}{2}\times 2^n$, the adversary can assume that it corresponds to the raw HLL cardinality estimate with high probability. In this case, no more useful information can be inferred about the bucket mappings, so the attacker has to reset the sketch to the empty state and start again with the non-discarded items. The adversary must pay attention not to shuffle the non-discarded items in between iterations, as the $b$ buckets to be filled are blindly determined by the first inserted items, and must stay the same throughout the attack. After several iterations, the adversary discards all items mapped to unwanted buckets, such that $b$ buckets are filled and the others left untouched.

The value of $b$ has to be picked smartly to keep the final cardinality estimate low, as the registers must hold values potentially very high. $b$ is going to be chosen such that even in the worst case scenario where all register values $M[j]$ hold the the maximal value of $l+1$, the Hit Counting algorithm is going to be used to produce the final estimation. So $b$ must satisfy:
\begin{align}
E:=\alpha_mm^2\cdot(\sum_{j=1}^m2^{-M[j]})^{-1} & \le \frac{5}{2}m \nonumber\\
\alpha_mm^2\cdot((m-b)\cdot2^{-0}+b\cdot2^{-(l+1)})^{-1} & \le \frac{5}{2}m \nonumber\\
b & \le m\cdot(1-\frac{2\alpha_m}{5})\cdot(1-2^{-(l+1)})^{-1} \label{eqn:1} 
\end{align}
with $m=2^n$. For example with $n = 8$ and $l = 24$, $b$ can take any value up to 184. The initial set of items to be filtered can be built by varying the flexible fields of the items $x$ the adversary is trying to insert.
%so that the adversary can ``hide'' $2^{t-n}$ elements while keeping the HLL cardinality estimate to 1.

\noindent\paragraph{Scenario S2} The adversary knows $h$ so it can adjust fields in the item $x$ it is trying to insert and keep only those for which the $\ell$ rightmost bits of $h(x)$ start with a one bit. In this case, the number of leading zero bits computed by the estimator of the destination bucket is always zero, so the estimator of this bucket is unlikely to be updated and the final averaged approximate cardinality is likely to be left unchanged.
For example, in the scenario of~\cite{portscanhll}, the adversary could change the source port of its IP packet until it is satisfied with its hash value. Given sufficient flexibility in the input $x$, the adversary can insert many items into the HLL sketch without increasing the cardinality markedly, by repeatedly sampling from the input domain.

When attacking an empty sketch, the adversary can keep this strategy and already mount a powerful attack, but can do better by following the approach presented under S1. Filtering items is easier in this scenario as the adversary can infer the bucket mapping of an item $x$ directly from the first $n$ bits of $h(x)$. A trivial approach for the adversary is to retain items mapped to buckets 1 to $b$. We argue why targeting $b$ buckets is a better approach in the impact analysis presented in Section~\ref{sec:impact}.

\noindent\paragraph{Scenario S4} The adversary can adopt the same strategy as for S2, but make the attack more efficient and reduce the sampling requirements with the additional information available in this scenario (namely, the count values per bucket in the HLL sketch). Therefore, it can allow some leading zero bits in the $\ell$ rightmost bits of $h(x)$ so long as this does not increase the current bucket estimate. In other words, if $h(x)$ is mapped to the $i$-th bucket with estimated cardinality $2^{M[i]}=2^{c_i + 1}$ (because a value with $c_i$ leading zeroes was previously observed for this bucket), then, instead of the strict restriction of having a one bit in the first position in the $\ell$ rightmost bits of $h(x)$, the adversary is satisfied when there is a one bit in any of the first $c_i+1$ positions in this substring. In case the adversary hits an empty bucket (for which the estimator is 0), it must skip that bucket completely. Under these conditions, the estimators in the buckets are never updated, and there is no effect on the cardinality estimate made by HLL.

A special case arises when the HLL is completely empty at the beginning of the attack. Here the attacker is forced to increase the estimate in at least one bucket. Once again, the adversary can choose to fill $b$ buckets following the strategy described in S2, leaving all the other buckets untouched.

\subsection{Impact}\label{sec:impact}
In this section we evaluate the impact of the above attacks on the HLL cardinality estimate. We assume that $h$ is a ``good" hash function, so that each of the bits of $h(x)$ can be regarded as being independent and uniformly random. We assume that the adversary has $t$ bits of flexibility in its choice of inputs.
